material and methods

### 2018-03-21 man/ms_1_material_methods.Rmd
change drouhgt and greenness description

The years 2005 and 2012 were chosen as drought years for this study because they have been documented as two of the worst drought years in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2015; @Trigo2013] and they were charecterized as extreme drought in our climatic data (Figure S1). Aunque sabemos que existen otros años de sequía extrema (e.g. 1995, 1981) [@VicenteSerrano2014], no se han elegido por una limitación metodológica, esto es, nuestro objetivo es analizar el efecto de la sequía y la resiliencia a los eventos de sequía utilizando información de satélite y de dendrocronología, y la información de satélite utilizada (ver mas adelante) solamente está disponible desde el año 2000 en adelante. We compare only these previous drought years in radial growth-time series, to place older drought events with those in 2005 and 2012 ... DUDA GUILLERMO

* ¿Decimos aquí algo así como?: Para los datos de radial-growth hemos computado la resiliencia en los 10 eventos mas severos (ver tabla S1)

@@

## Greenness data

As a proxy to primary growth, to characterize the vegetation greenness of *Q. pyrenaica* we used the *Enhanced Vegetation Index* (EVI) derived from MOD13Q1 product obtained by the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor [@Didan2015]. EVI and NDVI (*Normalized Difference Vegetation Index*) are the most common greenness vegetation indices. We used EVI instead of NDVI because EVI is more sensitive to changes in high-biomass areas (a serious shortcoming of NDVI); EVI reduces the influence of atmospheric conditions on vegetation index values, and EVI corrects for canopy background signals [@Huete2002; @Krapivin2015; @Cabello2012].

@@

After the filter out process, we built the annual EVI profile for each pixel and then computed the EVI's annual mean values and the EVI anomaly for each pixel for the period 2000 - 2016. EVI mean was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual (r = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88). We explored the temporal trends of EVI at a pixel scale. We computed the nonparametric Mann–Kendall test (Kendall, 1975; Mann, 1945), using the `trend` package in R [@Thorsten2017]. We set 0.05 the alpha level for the test.

>>>>

During the second half of the XX century, several extreme drought episodes were recorded for the Iberian Peninsula (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented as two of the worst drought years in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014 @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were charecterized as extreme drought in our climatic data (Appendices S1-S3). We focused on these two drought events because limitations about temporal availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events. The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Pascoa2017 (see Appendix S3)

@@

## Greenness data 
As a proxy of primary growth, we characterized the vegetation greenness of *Q. pyrenaica* by means of the *Enhanced Vegetation Index* (EVI) derived from MOD13Q1 product obtained by the *Moderate Resolution Imaging Spectroradiometer* (MODIS) sensor [@Didan2015]. Vegetation indices derived from remote sensing information are linked with primary productivity [@Ruimy1994] and they have been been widely used to derive descriptors of ecosystem functioning [*e.g* @AlcarazSegura2009]. These indices are linearly related to net primary productivity through the fraction of photosynthetically active radiation intercepted by green vegetation (*i.e.* vegetation greenness) [@Ruimy1994].

We used EVI instead of NDVI (*Normalized Difference Vegetation Index*; another of the most common greenness vegetation indices) because EVI is more sensitive to changes in high-biomass areas (a serious shortcoming of NDVI); EVI reduces the influence of atmospheric conditions on vegetation index values, and EVI corrects for canopy background signals [@Huete2002; @Krapivin2015; @Cabello2012].

@@

After the filter out process, we computed the annual mean EVI values as a surrogate of mean annual primary production, and the EVI anomaly for each pixel for the period 2000 - 2016. EVI mean was chosen since it is highly stable under the use of any filter [@Reyes2015] and because it showed high significant correlations with annual (r = 0.81) and seasonal EVI values ($r_{spring}$ = 0.76 and $r_{summer}$ = 0.88). We explored the temporal trends of EVI at a pixel scale. We computed the nonparametric Mann–Kendall test (Kendall, 1975; Mann, 1945), using the `trend` package in R [@Thorsten2017]. We set 0.05 the alpha level for the test.
